Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-08-15 _tex/paper2.tex
Built site for gh-pages

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Psychological Methods}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation},
  pdfauthor={Jose Manuel Rivera Espejo; Tine van Daal; Sven De Maeyer; Steven Gillis},
  pdfkeywords={causal inference, directed acyclic graphs, structural
causal models, bayesian statistical methods, thurstonian
model, comparative judgement, probability, statistical modeling},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Let's talk about Thurstone \& Co.: An information-theoretical
model for comparative judgments, and its statistical translation}
\author[1]{Jose Manuel Rivera Espejo%
\corref{cor1}%
}
 \ead{JoseManuel.RiveraEspejo@uantwerpen.be} 
\author[1]{Tine Daal%
%
}
 \ead{tine.vandaal@uantwerpen.be} 
\author[1]{Sven Maeyer%
%
}
 \ead{sven.demaeyer@uantwerpen.be} 
\author[2]{Steven Gillis%
%
}
 \ead{steven.gillis@uantwerpen.be}

@@

\begin{abstract}
This study revisits Thurstone's law of comparative judgment (CJ) by
addressing two key issues in the CJ literature. First, it critiques the
reliance on Thurstone's Case V assumptions and the Bradley-Terry-Luce
(BTL) model in the analysis of CJ data. Specifically, it argues that
while the assumptions of equal discriminal dispersions and zero
correlation simplify the trait measurement model, they may fail to
capture the complexity of some traits or stimuli, potentially leading to
unreliable and inaccurate trait estimates. Second, the study highlights
the disconnect between trait measurement and statistical inference in CJ
applications. It contends that while separating these processes
simplifies the analysis, this practice can also undermine the resulting
statistical inferences. To address these issues, this study extends
Thurstone's general form using a systematic, integrated approach based
on causal and Bayesian inference methods. This extension integrates CJ's
core theoretical principles alongside key assessment design features. It
then translates these elements into a statistical model for analyzing
dichotomous CJ data. Finally, the study emphasizes the relevance of this
extension for contemporary empirical CJ research, stressing the need for
CJ models tailored to the experiments and data assumptions. It also
encourages researchers across the social sciences to adopt more robust
and interpretable methodologies.
\end{abstract}

\begin{keyword}
    causal inference \sep directed acyclic graphs \sep structural causal
models \sep bayesian statistical methods \sep thurstonian
model \sep comparative judgement \sep probability \sep 
    statistical modeling
\end{keyword}
\end{frontmatter}

>>>>

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Psychological Methods}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation},
  pdfauthor={Jose Manuel Rivera Espejo; Tine van Daal; Sven De Maeyer; Steven Gillis},
  pdfkeywords={causal inference, directed acyclic graphs, structural
causal models, bayesian inference, thurstonian model, comparative
judgement, probability, statistical modeling},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Let's talk about Thurstone \& Co.: An information-theoretical
model for comparative judgments, and its statistical translation}
\author[1]{Jose Manuel Rivera Espejo%
\corref{cor1}%
}
 \ead{JoseManuel.RiveraEspejo@uantwerpen.be} 
\author[1]{Tine van Daal%
%
}
 \ead{tine.vandaal@uantwerpen.be} 
\author[1]{Sven De Maeyer%
%
}
 \ead{sven.demaeyer@uantwerpen.be} 
\author[2]{Steven Gillis%
%
}
 \ead{steven.gillis@uantwerpen.be}

@@

\begin{abstract}
This study revisits Thurstone's law of comparative judgment (CJ) by
addressing two key issues in the CJ literature. First, it critiques the
reliance on Thurstone's Case V assumptions and the Bradley-Terry-Luce
(BTL) model in the analysis of CJ data. Specifically, it argues that
while the assumptions of equal discriminal dispersions and zero
correlation simplify the trait measurement model, they may fail to
capture the complexity of some traits or stimuli, potentially leading to
unreliable and inaccurate trait estimates. Second, the study highlights
the disconnect between trait measurement and statistical inference in CJ
applications. It contends that while separating these processes
simplifies the analysis, this practice can also undermine the resulting
statistical inferences.

To address these issues, this study extends Thurstone's general form
using a systematic, integrated approach based on causal and Bayesian
inference methods. This extension integrates CJ's core theoretical
principles alongside key assessment design features. It then translates
these elements into a statistical model for analyzing dichotomous CJ
data.

Finally, the study emphasizes the relevance of this extension for
contemporary empirical CJ research, stressing the need for CJ models
tailored to the experiments and data assumptions. It also encourages
researchers across the social sciences to adopt more robust and
interpretable methodologies.
\end{abstract}

\begin{keyword}
    causal inference \sep directed acyclic graphs \sep structural causal
models \sep bayesian inference \sep thurstonian model \sep comparative
judgement \sep probability \sep 
    statistical modeling
\end{keyword}
\end{frontmatter}
